---
title: "DATA 607 Wk 1 Assignment"
author: "Daniel Moscoe"
date: "2/1/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

What regions in the United States are best equipped to deliver intensive care treatment to COVID-19 patients? This dataset compares the number of people at high-risk for requiring ICU treatment due to COVID-19 with the number of ICU beds in that region. In regions where the number of high-risk residents is large relative to ICU beds, there is greater danger that hospital resources will fail to meet the need for medical care. In regions where the number of high-risk residents is low relative to ICU beds, it is more likely that hospital resources will be adequate to the demand for ICU care during the pandemic. Click [here](https://fivethirtyeight.com/features/how-one-high-risk-community-in-rural-south-carolina-is-bracing-for-covid-19/) for the original article at FiveThirtyEight.

### Importing the data

```{r}
library(tidyverse)
data_on_github = "https://raw.githubusercontent.com/fivethirtyeight/data/master/covid-geography/mmsa-icu-beds.csv"

covid_geography <- read_csv(url(data_on_github))
```

### Selecting and renaming columns

```{r}

covid_geography <- covid_geography[, c(1, 2, 3, 5, 6, 7)]

covid_geography <- rename(covid_geography, c("region" = "MMSA", "percent_residents_at_risk" = "total_percent_at_risk", "residents_at_risk_per_ICU_bed" = "high_risk_per_ICU_bed", "ICU_beds" = "icu_beds", "residents_at_risk" = "total_at_risk"))

names(covid_geography)

```

### Visualizing residents at risk per ICU bed

```{r}
ggplot(data = covid_geography, mapping = aes(x = residents_at_risk_per_ICU_bed)) +
  geom_histogram()
```


### Conclusions

As noted in the FiveThirtyEight article, Hilton Head, SC is one of the regions in greatest danger. Only Manhattan, KS has a greater ratio of residents at risk to ICU beds. The authors may have chosen to focus on Hilton Head rather than Manhattan, KS because Hilton Head has roughly three times the number of residents at risk.

There is also significant variability across regions in the ratio of residents at risk to ICU beds, as can be seen in the histogram above.

Identifying which regions are most likely to face shortages of health care due to COVID-19 can help decision-makers move resources to where they're most needed.

To further investigate which regions of the US are at greatest risk of overwhelming their hospitals, it would be useful to have some measure of how flexible hospital resources are. The article described how Coastal Carolina Hospital has added ICU capacity, and studying how to train a more flexible workforce. Hospitals that are able to repurpose their space and staff are less likely to be overwhelmed, regardless of their initial ICU bed capacity.

### Questions

* Why does the residents_at_risk column, which purports to measure the number of residents at increased risk in a region, contain decimal values?

* Why couldn't I get the XML function htmlParse() to work? I was not able to reconstruct the example on p. 2 of _Automated Data Collection with R_. Using read_csv function is okay when using data posted to GitHub. But it doesn't have much use beyond that, I don't think.

* ?plyr::rename says that the old field name should come first, followed by the new field name. For example, renaming column "disp" to "displacement" is rename(mtcars, c("disp" = "displacement")). The cookbook-r.com link on Blackboard says the same thing. But my code only worked when the new field name was left of = and the old field name was right. Am I missing something?