---
title: "DATA607 Week 3 Assignment"
author: "Daniel Moscoe"
date: "2/15/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
```

*1. Using the 173 majors listed in fivethirtyeight.com's College Majors dataset, provide code that identifies the majors that contain either "DATA" or "STATISTICS".*

Import the dataset into R:

```{r}
college_majors_csv = "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"
college_majors <- read_csv(url(college_majors_csv))
```

Filter the data based on a regular expression in the Major column. The grepl function returns a vector of booleans based on whether the entry in Major matches the regular expression. Then filter returns rows from college_majors for which grepl returned True.

```{r}
ds_majors <- college_majors %>%
  filter(grepl("DATA|STATISTICS", Major))
ds_majors$Major
```

*2. Write code that transforms $\texttt{fruits}$, a vector of 14 character vectors, into a single string beginning $\texttt{c("}$.*

```{r}
fruits <- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

stem <- "c(" #begin the output string
for (fruit in fruits) { #add fruits and associated punctuation
  stem <- str_c(stem, '"', fruit, '", ')
}
stem <- str_c(substr(stem, 0, nchar(stem) - 2), ")") #replace final comma with closing parenthesis

print.noquote(stem) #display resulting string with escape characters hidden
```


*3. Describe, in words, what these expressions will match:*

$\texttt{(.)\1\1}$
This expression matches any string containing a single character occurring three times in a row (for example, eee). Since there are no English words that contain this pattern, the subset of $\texttt{words}$ matching this pattern is empty.

```{r}
str_subset(words, "(.)\\1\\1")
```

$\texttt{"(.)(.)\\\2\\\1"}$
This expression matches any string containing a symmetric sequence of 4 characters (for example, abba).

```{r}
str_subset(words, "(.)(.)\\2\\1")
```

$\texttt{(..)\1}$
This expression matches any string containing a repeated pair of letters (for example, abab).

```{r}
str_subset(words,"(..)\\1")
```

$\texttt{"(.).\\\1.\\\1"}$
This expression matches any string containing a sequence of 5 characters such that the first, third, and fifth character are the same (for example, abaca).

```{r}
str_subset(words, "(.).\\1.\\1")
```

$\texttt{"(.)(.)(.).*\\\3\\\2\\\1"}$
This expression matches any string containing a sequence of any three characters, followed later by that same sequence reversed (for example, abc...cba).

```{r}
str_subset(words, "(.)(.)(.).*\\3\\2\\1")
```

*4. Construct regular expressions to match words that:*  
*Start and end with the same character.*

$\texttt{^(.)}$ Begin the string with any character.  
$\texttt{.*}$ Follow with any characters.  
$\texttt{\1\$}$ End the string with a backreference to the character represented by the period enclosed in parentheses.

```{r}
str_subset(words,"^(.).*\\1$")
```

*Contain a repeated pair of letters (e.g., "church" contains "ch" repeated twice.)*

$\texttt{(..)}$ Identify a pair of characters.  
$\texttt{.*}$ Follow with any characters.  
$\texttt{\1}$ Follow with a backreference to the pair of characters identified earlier.

```{r}
str_subset(words, "(..).*\\1")
```

*Contain one letter repeated in at least three places (e.g., "eleven" contains three "e"s).*

$\texttt{(.)}$ Identify any character.  
$\texttt{.*}$ Follow with any characters.  
$\texttt{\1}$ Follow with a backreference to the character matched to (.).  
$\texttt{.*}$ Follow with any characters.  
$\texttt{\1}$ Follow with a backreference to the character matched to (.).
```{r}
str_subset(words,"(.).*\\1.*\\1")
```

*Questions*  
A programming language like Python is Turing complete because it can be used to express any computable function. Is there a sense in which regular expressions are "complete"? That is, can a regular expression be written to express any possible pattern in a string? Computers are built from primitive logic gates, like and, or, not. Is there some primitive set of operations for identifying patterns in strings?